---
title: "U.S. Inflation Analysis"
author: "Hannah May"
date: "March 2026"

format: 
  html:
    toc: true
    number-sections: true
    theme: cosmo
    code-fold: true
---

```{r}
packages <- c("fredr", "tidyverse", "devtools", "readxl", "lubridate")
suppressPackageStartupMessages({
  invisible(lapply(packages, library, character.only = TRUE))
})

fredr_set_key(Sys.getenv("FRED_API_KEY"))

start_date <- as.Date("2021-01-01")

fetch_fred_data <- function(series_id, start_date, end_date = Sys.Date()) {
  if (as.Date(end_date) < as.Date(start_date)) {
    stop("end_date must be after start_date")
  }
  fredr(
    series_id = series_id,
    observation_start = start_date,
    observation_end = as.Date(end_date)
  )
}
```
## Executive Summary

- Inflation has moderated from prior peaks but remains elevated relative to 2% target.
- Core inflation measures continue to show persistence, undicating underlying price pressures. 
- Real wage growth has recently exceeded productivity, suggesting demand-side inflation pressure
- Financial markets signal expectations of slowing growth, as reflected in an inverted yield curve.

## Inflation overview
```{r}
# CPI
df_cpi <- fetch_fred_data("CPIAUCSL", start_date) %>%
  select(date, value) %>%
  mutate(value = 100 * (value - lag(value, 12)) / lag(value, 12),
         series = "CPI")

# PCE
df_pce <- fetch_fred_data("PCEPI", start_date) %>%
  select(date, value) %>%
  mutate(value = 100 * (value - lag(value, 12)) / lag(value, 12),
         series = "PCE")

# Core CPI
df_core_cpi <- fetch_fred_data("CPILFESL", start_date) %>%
  select(date, value) %>%
  mutate(value = 100 * (value - lag(value, 12)) / lag(value, 12),
         series = "Core CPI")

# Core PCE
df_core_pce <- fetch_fred_data("PCEPILFE", start_date) %>%
  select(date, value) %>%
  mutate(value = 100 * (value - lag(value, 12)) / lag(value, 12),
         series = "Core PCE")

#Combine data
dat_inf <- bind_rows(df_cpi, df_pce, df_core_cpi, df_core_pce) %>%
  filter(!is.na(value), date >= as.Date("2022-01-01"))

#Plot
ggplot(dat_inf, aes(date, value, color = series)) +
  geom_line(aes(size = series)) +
  scale_size_manual(values = c("CPI"=1,"PCE"=1,"Core CPI"=1.4,"Core PCE"=1.4)) +
  guides(size = "none") +
  scale_color_manual(values = c("CPI"="black","Core CPI"="red","PCE"="darkblue","Core PCE"="lightblue")) +
  labs(title="Inflation Has Moderated but Remains Elevated",
       subtitle="Core inflation continues to show persistent underlying pressure",
       y="12-Month % Change", x="", color="",caption = "Source: BLS via FRED®") +
  scale_x_date(date_breaks="6 months", date_labels="%b %Y", expand = expansion(mult = c(0.02,0.05))) +
  theme_minimal() +
  theme(legend.position="top")

```

**Core inflation remains elevated relative to headline measures, indicating persistent underlying price pressures despite recent moderation.**

## Energy & Supply Shock

```{r}
df_oil <- fetch_fred_data("DCOILWTICO", start_date) %>%
  select(date, value) %>%
  filter(!is.na(value)) %>%
  filter(date >= as.Date("2021-01-01"))

ggplot(df_oil, aes(date, value)) +
  geom_line(color = "black", linewidth = 1.2) +
  
  geom_hline(
    yintercept = mean(df_oil$value, na.rm = TRUE),
    linetype = "dashed",
    color = "grey50"
  ) +
  
  labs(
    title = "Oil Prices Remain Volatile",
    subtitle = "Energy contributes to short-term inflation fluctuations",
    y = "Price ($)",
    x = "",
    caption = "Source: FRED®"
  ) +
  
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y",
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  
  theme_minimal()

```

**Energy prices remain volatile and contribute to short-term fluctuations in headline inflation.**

## Demand Pressure

```{r}
# Fetch raw CPI and earnings data
df_cpi_raw <- fetch_fred_data("CPIAUCSL", start_date) %>%
  select(date, cpi = value)

df_earn <- fetch_fred_data("CES0500000003", start_date) %>%
  select(date, earnings = value)

# Real wages
df_rw <- merge(df_cpi_raw, df_earn, by = "date") %>%
  mutate(
    real_wage = earnings / cpi,
    growth = 100 * (real_wage - lag(real_wage, 12)) / lag(real_wage, 12)
  ) %>%
  filter(!is.na(growth)) %>%
  filter(month(date) %in% c(1, 4, 7, 10)) %>%
  mutate(date = floor_date(date, "quarter")) %>%
  select(date, value = growth) %>%
  mutate(series = "Real Wage Growth")

# Productivity - values are ALREADY YoY % change, do not recompute
df_prod <- fetch_fred_data("PRS85006091", start_date) %>%
  select(date, value) %>%
  filter(!is.na(value)) %>%
  mutate(series = "Labor Productivity")

# Combine datasets
dat_dp <- bind_rows(df_rw, df_prod) %>%
  mutate(date = floor_date(date, "quarter")) %>%
  filter(date >= as.Date("2022-01-01"))

# Plot
ggplot(dat_dp, aes(date, value, color = series)) +
  
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  
  geom_line(linewidth = 1.4) +
  
  scale_color_manual(values = c(
    "Real Wage Growth" = "purple",
    "Labor Productivity" = "steelblue"
  )) +
  
  labs(
    title = "Real Wage Growth & Labor Productivity",
    subtitle = "Year-over-year percent change",
    x = "",
    y = "Year-over-Year % Change",
    color = "",
    caption = "Source: BLS via FRED®"
  ) +
  
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b %Y",
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  
  coord_cartesian(xlim = c(as.Date("2023-01-01"), NA)) +
  
  theme_minimal() +
  theme(legend.position = "top")
```

**Real wage growth has recovered sharply from deeply negative levels and is converging toward productivity growth, suggesting building demand-side inflationary pressure.**


## Market Signals

```{r}
df_yield <- fetch_fred_data("T10Y3M", start_date) %>%
  select(date, value) %>%
  filter(date >= as.Date("2022-01-01"))

ggplot(df_yield, aes(date,value)) +
  geom_ribbon(data=subset(df_yield,value<0),
              aes(ymin=value,ymax=0), fill="lightblue", alpha=0.3) +
  geom_line(color="steelblue", linewidth=1.4) +
  geom_hline(yintercept=0,color="black") +
  labs(title="Yield Curve Inversion Signals Slowing Growth",
       subtitle="Negative spread reflects expectations of moderating inflation",
       y="10Y - 3M Spread", x="", caption = "Source: BLS via FRED®") +
  scale_x_date(date_breaks="6 months", date_labels="%b %Y", expand = expansion(mult = c(0.02, 0.05))) +
  theme_minimal()


```

**The inverted yield curve suggests that financial markets anticipate slower economic growth and easing inflation pressures.**

## Outlook
While easing energy prices may continue to reduce headline inflation, renewed volatility in oil markets introduces upside risk. Converging wage and productivity growth, alongside elevated core inflation, suggest inflation is likely to decline gradually rather than rapidly.

Looking ahead, key indicators to monitor include:

- **Energy prices**, particularly oil, given their sensitivity to geopolitical developments  
- **Core inflation measures**, which reflect underlying price pressures  
- **Wage growth relative to productivity**, as a signal of demand-side inflation
- **Financial market indicators**, such as the yield curve, which provide insight into expectations for growth and inflation  

Overall, while inflation is expected to continue moderating, risks remain tilted toward a slower and uneven path back to target.

